Learning how to act: making good decisions with machine learning

### 2017-09-10 thesis/abstract.tex
fixes title pages

\documentclass[11pt,a4paper,twoside]{report}
\usepackage[table]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,fit,positioning}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib} %numbers
\usepackage{wrapfig}
\usepackage{centernot}

% this is included to get tables with colored rows to not draw over hlines and vlines
\usepackage{etoolbox}

\makeatletter
\patchcmd{\@classz}
  {\CT@row@color}
  {\oldCT@column@color}
  {}
  {}
\patchcmd{\@classz}
  {\CT@column@color}
  {\CT@row@color}
  {}
  {}
\patchcmd{\@classz}
  {\oldCT@column@color}
  {\CT@column@color}
  {}
  {}
\makeatother

\usepackage{amsthm}

\graphicspath{ {figures/} }
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
\DeclareUnicodeCharacter{00A0}{ }

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.3em,
           minimum height=2.3em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    latentrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

\newcommand{\KL}{\operatorname{KL}}

\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left(#1\right)}
\newcommand{\EE}{\mathbb E}
\newcommand{\EEa}{\EE_a}
\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
\newcommand{\actions}{\mathcal{A}}
\newcommand{\calA}{\mathcal A}
\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}
\newcommand{\R}{\mathbb R}
\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
\newcommand{\indep}[1]{\mathcal{I}\left(#1\right)}
\newcommand{\vb}[1]{\boldsymbol{#1}}
\newcommand{\card}[1]{\left\vert{#1}\right\vert}
\newcommand{\actionspace}{\mathcal{A}}

\newcommand{\V}[1]{\mathbb V\left[#1\right]}
\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}

\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}

\renewcommand{\P}[1]{\operatorname{P}\left(#1\right)}
\newcommand{\cf}[2]{{#1}^{#2}}
\newcommand{\cfb}[2]{{#1}_t^{#2}} % used to represent reward
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\regret}{\bar{R}_{T}} %psuedo-regret
\newcommand{\expectedregret}{\mathbb{E}R_{T}} % expected regret
\newcommand{\rawregret}{R_{T}} % plain regret
\newcommand{\simpleregret}{R_T}
\newcommand{\quotes}[1]{``#1''}

\theoremstyle{plain} 
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\let\temp\epsilon
\let\epsilon\varepsilon

\newcommand{\ci}{\mathrel{\perp\mspace{-10mu}\perp}}
\newcommand{\nci}{\centernot{\ci}}

\author{Finnian Lattimore}
\title{Learning how to act: making good decisions with machine learning}

\setcounter{secnumdepth}{3}

\begin{document}

%\def\ci{\perp\!\!\!\perp} % from Wikipedia

\begin{titlepage}
    \begin{center}
        \vspace*{1cm}

        \Huge
        {Learning how to act: making good decisions with machine learning}

        \vspace{2cm}
        \Large
        {Finnian Lattimore}

        \vfill
        \Large
        A thesis submitted for the degree of Doctor of Philosophy

         The Australian National University

        \vspace{1cm}

        September 2017

    \end{center}
\end{titlepage}

\epigraph{In vain the grave, with retrospective Eye,
Would from the apparent what conclude the why,
Infer the Motive from the Deed, and show
That what we chanced, was what we meant, to do.}{Alexander Pope}

\clearpage
\vspace*{\fill}
\begin{center}
\begin{minipage}{.9\textwidth}
Except where acknowledged in the customary manner, the material 
presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any 
university.

\vspace{40mm}  % vertical space

\hspace{80mm}\rule{40mm}{.15mm}\par   % horizontal space, line, start new line
\hspace{80mm} Finnian Lattimore\par
\hspace{80mm} September, 2017
\end{minipage}
\end{center}
\vfill % equivalent to \vspace{\fill}
\clearpage

\begin{abstract}

@@

\end{abstract}

%\chapter*{Acknowledgements}

\tableofcontents

\chapter{Introduction}

\section{Motivation}
Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure multiple sclerosis and would delaying teen-aged pregnancies improve the outcome for their children.

These are hard questions because they require more than identifying a pattern in data. Correlation is not causation. Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, despite the fact that the variables involved are susceptible to human intuition and understanding.

We now live in a world of data. Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and where the US Institute of Health is considering if all infants should be genetically sequenced at birth. Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by-product of some other purpose. It does not come from the controlled world of a randomised experiment.

The rise of big data sets and powerful computers has seen an explosion in the application of machine learning. From health care, to entertainment and self-driving cars, machine learning algorithms will transform many industries. It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory \citep{anderson2008end} and that we may be only a short step from \quotes{The Singularity}, where artificial intelligence exceeds our own and then grows exponentially.

However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within narrow problem settings. Getting them to generalise to even slightly different problems or data sets remains very challenging. Deciding how we should act or what policies we should implement requires predictions about how a system will behave if we change it. The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on. As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live in, it is critical to ensure we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.

\section{What is causality? }

The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined. This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends which variables are controlled for (or included in a regression). Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.

There are two ways in which an association detected in a data set may be non-causal. The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample. Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated. For example, based on data from the years 1999 to 2009, the age of Miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}. We would not expect this relationship to hold in a new sample of data. This form of spurious correlation also has serious repercussions. It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published. Consequently, the likelihood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}. This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}. This issue can be ameliorated by obtaining more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.

However, a strong association, observed in multiple independent studies may still not be causal. The correlation can arise because both variables are consequences of some other, unmeasured factor. For example, the reading ability of children under twelve is strongly correlated with their height, because older children are taller and can read better. However height is not a cause of reading ability because interventions to increase a child's height, for example by giving them growth hormones, would not be expected to improve their reading. Similarly, extra lessons in reading will not make a child grow taller. This problem is fundamentally different to the issue of spurious correlations arising by chance in finite data sets. Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to separate causation from correlation.

The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables. In this thesis, I take an interventionist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal. This viewpoint captures the types of questions that motivate this thesis. How can we change the way we do things to obtain better outcomes?

Causality is often liked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{what if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identifying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, that we describe in section \ref{sec:causal_models}. However, they state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on a given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are often relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "what if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see for example \citep{}.
\todo{reference a review paper on causal discovery here}

There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models. In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.

\section{What makes a problem causal?}

Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them. Machine learning techniques are being applied a huge range of problems, in both industry and academia. The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to. Which, if any, of these problems require causal inference?

\begin{itemize}
\item Speech recognition (for systems like Siri or Google Assistant)
\item Image classification
\item Forecasting the weather
\item Identifying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pneumonia.
\item Predicting who will re-offend on release from prison 
\item Customer churn modelling
\item Demand prediction for inventory control
\item Playing Go 
\end{itemize}

The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important consideration. In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.

Consider speech recognition. You say something, which causes sound waves, which are converted to a digital signal that Siri maps to words. Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution). A similar argument could be made for many applications of machine translation and image classification.

In image classification we do not particularly care about building a strong model for exactly how the thing that was photographed translates to an array of pixels, provided we can be fairly confident that the process will not change. If we develop a discriminative model that is highly accurate at classifying cats from dogs, we do not need to understand its internal workings (assuming we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).

What about forecasting the weather? If you are using a short term forecast to decide whether to pack an umbrella, causality can be ignored. Your decision will not affect if it actually rains. However, longer term climate forecasts might (theoretically) lead to action on emissions which would then change the weather system. For this we need a (causal) model that allows us to predict the outcome under various different interventions.

Identifying spam and automated essay marking both involve processing text to determine an underlying (complex) attribute such as its topic or quality. In both cases, there is inherent competition between the algorithm and the people generating the text. As a result, decisions made by the algorithm are likely to change the relationship between the features it relies on and the true label. Spammers and students will modify their writing in order to optimise their results. A standard supervised learning approach can only work if the resulting change in the mapping from features to label is sufficiently gradual. There are two key ways of ensuring this. The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm. The second is to use a model in which the features are related to the outcome in such a way that they cannot be manipulated independently.

This example also highlights a connection between causal models and transparency in machine learning. If we are using a non-causal model to make decisions affecting people, there will be a trade-off between the performance and transparency of the model; not because the requirement for transparency restricts us to simple models, but because revealing how the model works allows people to change their behaviour to game it.

What about predicting the risk of death in patients with pneumonia? Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective for serious cases, this appears to be straightforward prediction. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admission decisions. \citet{Caruana2015} found exactly this effect in a real data set. The model learned that people suffering asthma were \emph{less} likely to die from pneumonia. This was because doctors treated such patients very aggressively, thus actually lowering their risk. The issue is not with model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols. However, using it to decide how to \emph{change} these protocols could kill. The actual question of interest in this case is what happens to patients with characteristics $X$ they are assigned treatment according to decision rule (or policy) $\pi(X)$.

Predicting recidivism among paroled prisoners or customer churn also fit within the class of problems where the goal is to identify a group for which a problem will occur in order to target treatment (additional support and monitoring for people on parole, loyalty rewards to improve customer retention, hospitalisation for the severely ill). Predictive models can be applied to such problems where the most effective treatment is known for a given target group, and where deciding who to treat on the basis of the model predictions will not change the relationship between the features and outcome.

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc). Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.

Demand prediction seems like a relatively straightforward prediction problem. Models use features such as location, pricing, marketing, time of year and weather, to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However, depending on the way demand is measured, there is a potential data censoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The state of the board following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win. It must identify higher level features of the board state that are associated with winning.}. Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of a player selecting action $a$, given board state $s$. This allows the agent to estimate the likelihood of winning from a given starting state by simulating many times the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelihood an opponent will play a given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

The problem of playing go is causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach. The actions the agent has to explore are the sames ones as human players explored to generate the training data, and both have the same objective - to win the game. In addition, the state of the board encapsulates all the information relevant to selecting a move. These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.

Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predictive) inference. A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes. The second requirement is particularly difficult to satisfy when an algorithm is making important decisions affecting individual people. Think about problems like credit scoring and parole decisions. There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency. It is rare to build a model without any intent to make some kind of decision based on its results. Thus, I argue we should assume a causal model is required until we can justify otherwise.

\section{Observe or intervene: two very different approaches to causal problems}

As we have shown, problems involving causality are ubiquitous in many fields. As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning. Although the focus and terminology can differ substantially between fields these techniques, all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system. Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.

In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them. We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal. A particular strand of research within reinforcement learning are multi-armed bandit problems. They describe settings in which there are a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action. They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.

The approach of learning the outcome of an action by taking it plays a key role in advancing our knowledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the effect of an action from such observational data sets is the problem addressed by observational causal inference. Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system. This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another. The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.

Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process, and causal inference with problems where randomisation is not feasible, affordable or ethical. The similarities between the problems addressed by these techniques raises the question of whether there are problems best addressed by a combination of these approaches, and if so, how they can be combined.

\section{This thesis and its contributions}

I view causal problems as one of the greatest current challenges for machine learning. They incorporate a large set of problems of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence. For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry, we need to understand how to leverage our improved approaches to prediction to tackle causal problems.

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty. 
I make a number of additional connective contributions that are not encompassed by my work on causal bandit problems. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift. I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

\paragraph{Thesis overview} This thesis is divided into three key chapters, learning from observational data, learning from interventional data and unifying the approaches. Chapter \ref{chap:causality} covers learning to act from observational data. In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems. Section \ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \ref{sec:defining-causal-effects} and \ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation. Section \ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.

Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems. In \S\ref{sec:multi-armed-bandits}, I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}). I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}. In \S\ref{sec:off-policy-evaluation}, I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of causal effect estimation from observational data. Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems: a framework that unifies causal graphical models with multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

\chapter{Learning from observational data}
\label{chap:causality}

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system from observing it. This is the problem of observational causal inference.

\section{Causal models}
\label{sec:causal_models}

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention. Mapping properties from one system to another requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate a system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised \citep{Pearl2000}) framework and associated set of methods for estimating causal effects. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains spit between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.

In the next sections, we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ. By describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them. However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers. In order to demonstrate the notation and formalisms each framework provides, we will use them to describe the following simple examples.

\vspace*{.3cm}
\begin{example}
\label{exm:ranomized_experiment}
Suppose a pharmaceutical company wants to assess the effectiveness of a new drug on recovery from a given illness. This is typically tested by taking a large group of representative patients and randomly assigning half of them to a treatment group (who receive the drug) and the other half to a control group (who receive a placebo). The goal is to determine the clinical impacts of the drug by comparing the differences between the outcomes for the two groups (in this case, simplified to only two outcomes - recovery or non-recovery). We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome. 
\end{example}

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds. We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
\end{example}

\subsection{Causal Bayesian networks}

Causal Bayesian networks are an extension of Bayesian networks. A Bayesian network is a graphical way of representing how a distribution factorises. Any joint probability distribution can be factorised into a product of conditional probabilities. There are multiple valid factorisations, corresponding to permutations of variable ordering.

\eqn{
\label{eqn:cbn:joint_dist}
P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
}

We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}. If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network. There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

>>>>

\chapter*{Abstract}

@@
